Efficiently Handling Guards in the low level design of RPython's tracing JIT

### 2012-07-03 talk/vmil2012/paper.tex
get started with some ideas for the RPython section

% compressing itemize env, in case we need it
\newenvironment{zitemize}% zero - line spacing itemize environment
   {\begin{list}{--}{
   \setlength{\itemsep}{0 pt}
   \setlength{\parsep}{0 pt}
   \setlength{\topsep} {0 pt} }}% the end stuff
   {\end{list}}

@@

\title{Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages}

\authorinfo{Carl Friedrich Bolz$^a$ \and Antonio Cuni$^{a, c}$ \and Maciej Fijałkowski$^b$ \and Michael Leuschel$^a$ \and \\
            Samuele Pedroni$^c$ \and Armin Rigo$^a$}
           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany

            $^b$merlinux GmbH, Hildesheim, Germany

            $^c$Open End, Göteborg, Sweden
           }
           {cfbolz@gmx.de \and anto.cuni@gmail.com \and fijal@merlinux.eu \and
           leuschel@cs.uni-duesseldorf.de \and samuele.pedroni@gmail.com \and arigo@tunes.org}

\conferenceinfo{ICOOOLPS'11,}{July 26, 2011, Lancaster, UK.}
\CopyrightYear{2011}
\crdata{978-1-4503-0894-6/11/07}

@@

\keywords{Tracing JIT, Runtime Feedback, Interpreter, Code Generation, Meta-Programming}

@@

Meta-tracing JIT compilers can be applied to a variety of different
languages without explicitly encoding language semantics into the compiler. So
far, they lacked a way to give the language implementor control over runtime
feedback. This restricted their performance. In this paper we describe the
mechanisms in PyPy's meta-tracing JIT that can be used to control
runtime feedback in language-specific ways. These mechanisms are flexible
enough to express classical VM techniques such as maps and runtime type feedback.

@@

One of the hardest parts of implementing an object-oriented dynamic language well is to
optimize its object model. This is made harder by the complexity of the core
object semantics of many recent languages such as Python, JavaScript or Ruby.
For them, even implementing just an interpreter is already a difficult
task. Implementing these languages efficiently with a just-in-time compiler (JIT) is
extremely challenging, because of their many corner-cases.

It has long been an objective of the partial evaluation community to
automatically produce compilers from interpreters. There has been a
renaissance of this idea around the approach of tracing just-in-time
compilers. A number of projects have attempted this approach. SPUR \cite{bebenita_spur:_2010} is
a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython \cite{davide_ancona_rpython:_2007} (a restricted
subset of Python). This JIT is then used to trace a number of languages
implementations written in RPython. A number of other experiments in this
directions were done, such as an interpreter for Lua in JavaScript, which is run
on and optimized with a tracing JIT for JavaScript
\cite{yermolovich_optimization_2009}.

These projects have in common that they work one meta-level down, providing a tracing JIT for the
language used to implement the dynamic language, and not for the dynamic language itself.
The tracing JIT will then trace through the object model of the dynamic
language implementation. This makes the object model transparent to the tracer
and its optimizations. Therefore the semantics of the dynamic language does not
have to be replicated in a JIT. We call this approach \emph{meta-tracing}.
Another commonality of these approaches is that they allow some annotations (or
hints) in the dynamic language implementation to guide the meta-tracer. This
makes the process not completely automatic but can give good speedups over
bare meta-tracing.

In this paper we present two of these hints that are extensively used in the
PyPy project to improve the performance of its Python interpreter, particularly
of the object model.

Conceptually, the significant speed-ups that can be achieved with
dynamic compilation depend on feeding into compilation values observed at
runtime and exploiting them. In particular, if
there are values which vary very slowly, it is possible to compile multiple
specialized versions of the same code, one for each actual value.  To exploit
the runtime feedback, the implementation code and data structures need to be
structured so that many such slow-varying values are at hand. The hints that
we present precisely allow us to implement such feedback and exploitation in a
meta-tracing context.

Together these hints can be used to express many classic implementation
techniques used for object models of dynamic languages, such runtime type
feedback and maps.

The contributions of this paper are:
\begin{itemize}
 \item A hint to turn arbitrary variables into constants in the trace by
 feeding back runtime information into compilation.
 \item A way to annotate operations which the constant folding
 optimization then recognizes and exploits.
 \item A worked-out example of a simple object model of a dynamic language and
 how it can be improved using these hints.
 \item This example also exemplifies general techniques for refactoring code to
 expose constant folding opportunities of likely runtime constants.
\end{itemize}

The paper is structured as follows: Section~\ref{sec:Background} gives an
introduction to the PyPy project and meta-tracing and presents an example of a
tiny dynamic language object model. Section~\ref{sec:hints} presents the hints,
what they do and how they are applied. Section~\ref{sec:fastobjmodel} shows how
the hints are applied to the tiny object model and Section~\ref{sec:evaluation}
presents benchmarks.

@@

\subsection{The PyPy Project}
\label{sub:pypy}

The PyPy project \cite{armin_rigo_pypys_2006} strives to be an environment where
complex dynamic languages can be implemented efficiently. The approach taken
when implementing a language with PyPy is to write an interpreter for the language
in \emph{RPython}. RPython is a restricted subset of Python chosen in such a way
that it is possible to perform type inference on it. The interpreters in RPython
can therefore be translated to efficient C code.

A number of languages have been implemented with PyPy, most importantly a full
Python implementation, but also a Prolog interpreter
\cite{carl_friedrich_bolz_towards_2010} and some less mature experiments.

The translation of the interpreter to C code adds a number of implementation details into the
final executable that are not present in the interpreter implementation, such as
a garbage collector. The interpreter can therefore be kept free from low-level
implementation details. Another aspect of the final VM that is added
semi-automatically to the generated VM is a tracing JIT compiler.

The advantage of this approach is that writing an interpreter is much easier
and less error prone than manually writing a JIT compiler.  Similarly, writing
in a high level language such as RPython is easier than writing in C.

We call the code that runs on top of a VM implemented with PyPy the
\emph{user code} or \emph{user program}.

%___________________________________________________________________________
\subsection{PyPy's Meta-Tracing JIT Compilers}
\label{sub:tracing}

A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs were popularized by the Dynamo project, which used the technique for dynamic
machine code optimization \cite{bala_dynamo:_2000}. Later they were used to implement
a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
JavaScript \cite{gal_trace-based_2009}.

A tracing JIT works by recording traces of concrete execution paths through the
program. Those
traces are linear lists of operations, which are optimized and then
get turned into machine code. This recording automatically inlines functions:
when a function call is encountered the operations of the called functions are
simply put into the trace of the caller too. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{andreas_gal_incremental_2006}.

Because the traces always correspond to a concrete execution they cannot
contain any control flow splits. Therefore they encode the control flow
decisions needed to stay on the trace with the help of \emph{guards}. Those are
operations that check that the assumptions are still true when the compiled trace is later executed with different values.

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used; only the most frequently executed paths through the user
program are traced and turned into machine code. The interpreter is also used when a guard
fails to continue the execution from the failing guard.

Since PyPy wants to be a general framework, we want to reuse our tracer for
different languages, which makes classical tracers inapplicable, because they
encode language semantics.
Therefore PyPy's JIT is a \emph{meta-tracer} \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces do not contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

Tracing through the execution of an interpreter has many advantages. It makes
the tracer, its optimizers and backends reusable for a variety of languages. The
language semantics do not need to be encoded into the JIT. Instead the tracer
just picks them up from the interpreter. This also means that the JIT by
construction supports the full language as correctly as the interpreter.

While the operations in a trace are those of the interpreter, the loops that are
traced by the tracer are the loops in the
user program. To achieve this the tracer stops tracing after one iteration of
the loop in the user function that is being considered; at this point, it probably
traced many iterations of the interpreter main loop.

\begin{figure}
\includegraphics[scale=0.45]{figures/trace-levels}
\caption{The levels involved in tracing}
\label{fig:trace-levels}
\end{figure}

Figure~\ref{fig:trace-levels} shows a diagram of the process. On the left are
the levels of execution. The CPU executes the binary of
PyPy's Python interpreter, which consists of RPython functions that have been
compiled first to C, then to machine code. The interpreter runs a Python program written by a
programmer (the user). If the tracer is used, it traces operations on the level
of the interpreter. However, the extent of the trace is determined by the loops
in the user program.

\subsection{Optimizing Traces}
\label{sub:optimizing}

Before sending the trace to the backend to produce actual machine code, it is
optimized.  The optimizer applies a number of techniques to remove or simplify
the operations in the trace. Most of these are well known compiler optimization
techniques, with the difference that it is easier to apply them in a tracing
JIT because it only has to deal with linear traces.  Among the techniques are
constant folding, common subexpression elimination, allocation removal
\cite{bolz_allocation_2011}, store/load propagation, loop invariant code
motion.

In some places it turns out that if the interpreter author rewrites some parts
of the interpreter with these optimizations in mind the traces that are produced
by the optimizer can be vastly improved.

\subsection{Running Example}
\label{sub:running}
\anto{this example is not referenced until section \ref{sec:fastobjmodel}: I
  would put it just before it}

As the running example of this paper we will use a very simple and bare-bones
object model that just supports classes and instances, without any
inheritance or other advanced features. In the model classes contain methods.
Instances have a class. Instances have their own attributes (or fields). When looking up an
attribute of an instance, the instance's attributes are searched. If the
attribute is not found there, the class' methods are searched.

\begin{figure}
\input{code/interpreter-slow.tex}
\caption{Original Version of a Simple Object Model}
\label{fig:interpreter-slow}
\end{figure}

To implement this object model, we use the RPython code in
Figure~\ref{fig:interpreter-slow} as part of the interpreter source code.
In this straightforward implementation the methods and attributes are just
stored in dictionaries (hash maps) on the classes and instances, respectively.
While this object model is very
simple it already contains most hard parts of Python's object model. Both
instances and classes can have arbitrary fields, and they are changeable at
any time.  Moreover, instances can change their class after they have been
created.

When using this object model in
an interpreter, a large amount of time will be spent doing lookups in these
dictionaries.
Let us assume we trace through code that sums three attributes, such as:
\anto{I still think it's a bit weird to call them ``methods'' and then use
  them as attributes in the example}

{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
inst.getattr("a") + inst.getattr("b") + inst.getattr("c")
\end{lstlisting}
}

\begin{figure}
\input{code/trace1.tex}
\caption{Trace Through the Object Model}
\label{fig:trace1}
\end{figure}

\cfbolz{should we show the code that would create the inst use in tracing?}

The trace would look like in Figure~\ref{fig:trace1}. In this example, the
attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The line
numbers in the trace correspond to the line numbers in
Figure~\ref{fig:interpreter-slow} where the traced operations come from. The
trace is in SSA form. Note how all the guards in trace correspond to a
condition in the original code. The trace contains
five calls to \texttt{dict.get}, which is slow. To make the language efficient
using a tracing JIT, we need to find a way to get rid of these dictionary
lookups. How to achieve this will be the topic of
Section~\ref{sec:fastobjmodel}.

% subsection Running Example (end)

% section Background (end)
%___________________________________________________________________________

\section{Hints for Controlling Optimization}
\label{sec:hints}

In this section we will describe two hints that allow the
interpreter author to increase the optimization opportunities for constant
folding.
If applied correctly these techniques can give really big speedups by
pre-computing parts of what happens at runtime. On the other
hand, if applied incorrectly they might lead to code bloat, thus making the
resulting program actually slower. Note that these hints are \emph{never} put into the user program, only into the interpreter.

For constant folding to work, two conditions need to be met: the arguments of
an operation actually need to all be constant, i.e. statically known by the
optimizer and the operation needs to be \emph{constant-foldable}, i.e. always
yield the same result given the same arguments.
There is one kind of hint for both of these conditions.

\subsection{Where Do All the Constants Come From?}

It is worth clarifying what a ``constant'' is in this context.  A variable of
the trace is said to be constant if its value is statically known by the
optimizer.
The simplest example of constants are literal values, such as \texttt{1}.
However, the optimizer can statically know the value of a variable even if it
is not a constant in the original source code. For example, consider the
following fragment of RPython code on the left. If the fragment is traced with
\texttt{x} being \texttt{4}, the trace on the right is produced:

\begin{minipage}[b]{0.5\linewidth}
\centering
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
if x == 4:
    y = y + x
\end{lstlisting}
}
\end{minipage}
\vline
\hspace{0.5cm}
\begin{minipage}[b]{0.5\linewidth}
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$y_2$ = $y_1$ + $x_1$
\end{lstlisting}
}
\end{minipage}

A guard is a runtime check. The above trace will run to completion when $x_1$
\texttt{== 4}. If the check fails, execution of the trace is stopped and the
interpreter continues to run. Therefore, the value of $x_1$ is statically known
to be \texttt{4} after the guard.

There are cases in which it is useful to turn an arbitrary variable
into a constant value. This process is called \emph{promotion} and it is an old idea
in partial evaluation (it's called ``The Trick''  \cite{jones_partial_1993}
there). The technique is substantially more powerful in a JIT compiler than in
the static setting of classic partial evaluation.

Promotion is essentially a tool for trace specialization. There are places in
the interpreter where it would open a lot of optimization opportunities if a
variable were constant, even though it could have different values in
practice. In such a place, promotion can be used. The typical reason to do that
is if there is a lot of computation depending on the value of one variable.

Let us make this more concrete. If we trace a call to the function (written in
RPython) on the left, we get the trace on the right:

\begin{minipage}[b]{0.5\linewidth}
\centering
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    z = x * 2 + 1
    return z + y
\end{lstlisting}
}
\end{minipage}
\vline
\hspace{0.5cm}
\begin{minipage}[b]{0.5\linewidth}
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}
}
\end{minipage}

Observe how the first two operations could be constant-folded if the value of
$x_1$ were known. Let us assume that the value of \texttt{x} in the RPython code can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:

\begin{minipage}[b]{0.5\linewidth}
\centering
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    promote(x)
    z = x * 2 + 1
    return z + y
\end{lstlisting}
}
\end{minipage}
\vline
\hspace{0.5cm}
\begin{minipage}[b]{0.5\linewidth}
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}
}
\end{minipage}

The hint indicates that \texttt{x} is likely a runtime constant and the JIT
should try to perform runtime specialization on it
in the code that follows. When just running
the code, the \texttt{promote} function has no
effect. When tracing, some extra work
is done. Let us assume that this changed function is traced with
the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
operation at the beginning.

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the runtime value of \texttt{x} as it was during tracing, which can
then be exploited by the compiler. The introduced
guard specializes the trace, because it only works if the value of $x_1$ is
\texttt{4}. From the point of view of the
optimizer, this guard is not different frome the one produced by the \texttt{if}
statement in the first example. After the guard, it can be assumed that $x_1$
is equal to \texttt{4}, meaning that the optimizer will turn this trace into:

{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_2$ = 9 + $y_1$
return($v_2$)
\end{lstlisting}
}

Notice how the first two arithmetic operations were constant folded. The hope is
that the guard is executed quicker than the multiplication and the addition that
was now optimized away.

If this trace is executed with values of $x_1$ other than \texttt{4}, the guard will
fail, and execution will continue in the interpreter. If the guard fails often
enough, a new trace will be started from the guard. This other trace will
capture a different value of $x_1$. If it is e.g. \texttt{2}, then the optimized
trace looks like this:

{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 2)
$v_2$ = 5 + $y_1$
return($v_2$)
\end{lstlisting}
}

This new trace will be attached to the guard instruction of the first trace. If
$x_1$ takes on even more values, a new trace will eventually be made for all of them,
linking them into a chain. This is clearly not desirable, so we should promote
only variables that do not vary much. However, adding a promotion hint will never produce wrong
results. It might just lead to too much machine code being generated.

Promoting integers, as in the examples above, is not used that often.
However, the internals of dynamic language interpreters often
have values that are variable but vary little in the context of parts of a user
program. An example would be the types of variables in a user function, which
rarely change in a dynamic language in practice (even though they could). In the
interpreter, these user-level types are values. Thus promoting them will lead
to type-specialization on the level of the user program.
Section~\ref{sec:fastobjmodel} will present a complete example of how this
works.

\subsection{Declaring New Foldable Operations}

In the previous section we saw a way to turn arbitrary variables into constants. All
foldable operations on these constants can be constant-folded. This works well for
constant folding of primitive types, e.g. integers. Unfortunately, in the context of an
interpreter for a dynamic
language, most operations actually manipulate objects, not primitive types. The
operations on objects are often not foldable and might even have side-effects. If
one reads a field out of a constant reference to an object this cannot
necessarily be folded away because the object can be mutated. Therefore, another
hint is needed.

This hint can be used to mark functions as \emph{trace-elidable}. A function is
termed trace-elidable if, during the execution of the program,
successive calls to the function with identical arguments always return the
same result. In addition the function needs to have no side effects or
idempotent side effects\footnote{This property
is less strict than that of a pure function, because it is only about actual
calls during execution. All pure functions are trace-elidable though.}.
From this definition follows that a call to a trace-elidable function with
constant arguments in a trace can be replaced with the result of the call seen during tracing.

As an example, take the class on the left. Tracing the call \texttt{a.f(10)} of
some instance of \texttt{A} yields the trace on the right (note how the call to
\texttt{c} is inlined):

\begin{minipage}[t]{0.6\linewidth}
\centering
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

    def f(self, val):
        self.y = self.c() + val

    def c(self):
        return self.x * 2 + 1
\end{lstlisting}
\end{minipage}
\vline
\hspace{0.5cm}
\begin{minipage}[t]{0.4\linewidth}
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$x_1$ = $a_1$.x
$v_1$ = $x_1$ * 2
$v_2$ = $v_1$ + 1
$v_3$ = $v_2$ + $val_1$
$a_1$.y = $v_3$
\end{lstlisting}
\end{minipage}

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{c}
is a trace-elidable function. To communicate this, there is an \texttt{@elidable} decorator.
If the code in \texttt{c} should be constant-folded away, we would change the
class as follows:

\begin{minipage}[t]{0.6\linewidth}
\centering
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

    def f(self, val):
        promote(self)
        self.y = self.c() + val

    @elidable
    def c(self):
        return self.x * 2 + 1
\end{lstlisting}
\end{minipage}
\vline
\hspace{0.4cm}
\begin{minipage}[t]{0.4\linewidth}
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 
      0xb73984a8)
$v_1$ = A.c($a_1$)
$v_2$ = $v_1$ + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
\end{minipage}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{A.c} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{A.c} method is marked as trace-elidable, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this (assuming that the \texttt{x} field's value is \texttt{4}):
%
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
}

On the one hand, the \texttt{@elidable} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives the interpreter author ample opportunity to introduce bugs. If a
function is annotated to be trace-elidable, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation\footnote{The most common use case of the \texttt{@elidable}
annotation is indeed to declare the immutability of fields. Because it is so
common, we have special syntactic sugar for it.}. We hope to introduce a
debugging mode which would (slowly) check whether the annotation is applied
incorrectly to mitigate this problem.

@@

\section{Putting It All Together}
\label{sec:fastobjmodel}

In this section we describe how the simple object model from
Section~\ref{sub:running} can be made efficient using the hints described in the
previous section. The object model there is typical for many current
dynamic languages (such as Python, Ruby and JavaScript) as it relies heavily on
hash-maps to implement its objects.

\subsection{Making Instance Attributes Faster Using Maps}

The first step in making \texttt{getattr} faster in our object model is to optimize
away the dictionary lookups on the instances. The hints of the previous section
do not seem to help with the current object model. There is
no trace-elidable function to be seen, and the instance is not a candidate for promotion,
because there tend to be many instances.

This is a common problem when trying to apply hints. Often, the interpreter
needs a small rewrite to expose the trace-elidable functions and nearly-constant objects
that are implicitly there. In the case of instance fields this rewrite is not
entirely obvious. The basic idea is as follows. In theory instances can have
arbitrary fields. In practice however many instances share their layout (i.e.
their set of keys) with many other instances.

Therefore it makes sense to factor the layout information out of the instance
implementation into a shared object, called the \emph{map}. Maps are a well-known
technique to efficiently implement instances and come from the SELF project
\cite{chambers_efficient_1989}. They are also used by many JavaScript implementations such as V8.
The rewritten \texttt{Instance} class using maps can be seen in
Figure~\ref{fig:maps}.

\begin{figure}
\input{code/map.tex}
\caption{Simple Object Model With Maps}
\label{fig:maps}
\end{figure}

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. Maps are shared between
different instances, therefore they have to be immutable, which means
that their \texttt{getindex} method is a trace-elidable function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map. This function is also trace-elidable,
because it caches all new instances of \texttt{Map} that it creates, to make
sure that objects with the same layout have the same map, which makes its side
effects idempotent. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

With this adapted instance implementation, the trace we saw in Section~\ref{sub:running} changes to
that of Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
memory address of the \texttt{Map} instance that has been promoted. Operations
that can be optimized away are grayed out, their results will be replaced with
fixed values by the constant folding.

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
a trace-elidable function and they have constant arguments. That means that $index_{1/2/3}$
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. This trace is already much better than
the original one. Now we are down from five dictionary lookups to just two.

The technique to make instance lookups faster is applicable in more general
cases. A more abstract view of maps is that of splitting a data-structure into
an immutable part (\eg the map) and a part that changes (\eg the storage list).
All the computation on the immutable part is trace-elidable so that only the
manipulation of the quick-changing part remains in the trace after
optimization.

\begin{figure}
\input{code/trace2.tex}
\caption{Unoptimized Trace After the Introduction of Maps}
\label{fig:trace2}
\end{figure}

\subsection{Versioning of Classes}

Above we assumed that the total number of
different instance layouts is small compared to the number of instances. For classes we
will make an even stronger assumption. We simply assume that it is rare for
classes to change at all. This is not always reasonable (sometimes classes contain
counters or similar things) but for this simple example it is good
enough.\footnote{There is a more complex variant of the presented technique that can
accommodate quick-changing class fields a lot better.}

What we would really like that the \texttt{Class.find\_method} method is trace-elidable.
But it cannot be, because it is always possible to change the class itself.
Every time the class changes, \texttt{find\_method} can potentially return a
new value.

Therefore, we give every class a version object, which is changed every time a
class gets changed (i.e., the \texttt{methods} dictionary changes).
This means that the result of calls to \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a trace-elidable operation.  To help
the JIT to detect this case, we factor it out in a helper method \texttt{\_find\_method} which is
marked as \texttt{@elidable}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

\begin{figure}
\input{code/version.tex}
\caption{Versioning of Classes}
\label{fig:version}
\end{figure}

What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
argument but it does not use it at all. Its only purpose is to make the call
trace-elidable, because when the version object changes, the result of the call might be
different from the previous one.

\begin{figure}
\input{code/trace4.tex}
\caption{Unoptimized Trace After Introduction of Versioned Classes}
\label{fig:trace4}
\end{figure}

The trace with this new class implementation can be seen in
Figure~\ref{fig:trace4}.
The calls to \texttt{Class.\_find\_method} can now be optimized away, also the
promotion of the class and the version, except for the first one. The final
optimized trace can be seen in Figure~\ref{fig:trace5}.

\begin{figure}
\input{code/trace5.tex}
\caption{Optimized Trace After Introduction of Versioned Classes}
\label{fig:trace5}
\end{figure}

The index \texttt{0} that is used to read out of the \texttt{storage} list is the result
of the constant-folded \texttt{getindex} call.
The constants \texttt{41} and \texttt{17} are the results of the folding of the
\texttt{\_find\_method} calls. This final trace is now very good. It no longer performs any
dictionary lookups. Instead it contains several guards. The first guard
checks that the map is still the same. This guard will fail if the same
code is executed with an instance that has another layout. The second guard
checks that the class of \texttt{inst} is still the same. It will fail if the trace is
executed with an instance of another class. The third guard checks that the
class did not change since the trace was produced. It will fail if somebody
calls the \texttt{write\_method} method on the class.

\subsection{Real-World Considerations}

The techniques used above for the simple object model are used for the object
model of PyPy's Python interpreter too. Since Python's object model is
considerably more complex, some additional work needs to be done.

The first problem that needs to be solved is that Python supports (multiple)
inheritance. Therefore looking up a method in a class needs to consider all the
classes in the
whole method resolution order. This makes the versioning of classes more
complex. If a class is changed its version changes. At the same time, the
versions of all the classes inheriting from it need to be changed as well,
recursively. This makes class changes expensive, but they should be rare.  On the
other hand, a method lookup in a complex class hierarchy is as optimized in the
trace as in our simple object model above.

Another optimization is that in practice the shape of an instance is correlated
with its class. In our code above, we allow both to vary independently.
In PyPy's Python interpreter we store the class of an instance on its map. This
means that we get one fewer promotion and thus one fewer
guard in the trace, because the class doesn't need to be promoted after the
map has been.

%\subsection{More General Patterns}
%
%
%Similarly, versions can be used to constant-fold arbitrary functions of large data
%structures. The version needs to be updated carefully every time the result of
%this function can change. Therefore this is useful only if the data structure is
%expected to change slowly.

@@

For space reasons we cannot perform a full evaluation here, but still want to
present some benchmark numbers. We chose to present just some benchmarks: The
templating engine of the Django web
framework\footnote{\texttt{http://www.djangoproject.com/}}; a Monte-Carlo Go
AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/
disco-elegant-python-go-player.html}}; a BZ2 decoder; a port of the classical
Richards benchmark to Python; a Python version of the Telco decimal
benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a
pure Python decimal floating point implementation. The results we see in these
benchmarks seem to repeat themselves in other benchmarks using object-oriented
code; for purely numerical algorithms the speedups introduced by the techniques
in this paper are much smaller because they are already fast.

The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
2.6.35. We compared the performance of two Python implementations on the
benchmarks. As a baseline, we used the standard Python implementation in C,
CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. We compare it against two versions of PyPy's Python interpreter,
both of them with JIT enabled. The PyPy baseline does not enable maps or type
version, the full JIT enables both.

All benchmarks were run 50 times in the same process, to give the JIT time to
produce machine code. The arithmetic mean of the times of the last 30 runs were
used as the result. The errors were computed using a confidence interval with a
95\% confidence level \cite{georges_statistically_2007}. The results are
reported in Figure~\ref{fig:times}, together with the same numbers normalized to
those of the full JIT.

The optimizations give a speedup between 80\% and almost 20 times. The Richards benchmark
is a particularly good case for the optimizations as it makes heavy uses of
object-oriented features. Pyflate uses mostly imperative code, so does not
benefit as much. Together with the optimization, PyPy outperforms CPython in
all benchmarks, which is not surprising because CPython is a simple
bytecode-based interpreter.

\begin{figure}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|}
\hline
	&CPython	&JIT baseline	&JIT full\\
\hline
django[ms]	&988.67 $\pm$ 0.49	&405.62 $\pm$ 4.80	&149.31 $\pm$ 1.37\\
	&6.62 $\times$	&2.72 $\times$	&1.00 $\times$\\
\hline
go[ms]	&947.43 $\pm$ 1.30	&525.53 $\pm$ 7.67	&174.32 $\pm$ 7.78\\
	&5.44 $\times$	&3.01 $\times$	&1.00 $\times$\\
\hline
pyflate[ms]	&3209.20 $\pm$ 3.65	&2884.26 $\pm$ 21.11	&1585.48 $\pm$ 5.22\\
	&2.02 $\times$	&1.82 $\times$	&1.00 $\times$\\
\hline
richards[ms]	&357.79 $\pm$ 1.32	&421.87 $\pm$ 0.48	&17.89 $\pm$ 1.15\\
	&20.00 $\times$	&23.58 $\times$	&1.00 $\times$\\
\hline
telco[ms]	&1209.67 $\pm$ 2.20	&738.18 $\pm$ 3.29	&153.48 $\pm$ 1.86\\
	&7.88 $\times$	&4.81 $\times$	&1.00 $\times$\\
\hline
\end{tabular}
}
\end{center}
\caption{Benchmark Results}
\label{fig:times}
\end{figure}

@@

In this paper we presented two hints that can be used in the source code of an
interpreter written with PyPy. They give control over runtime feedback and
optimization to the language implementor. They are expressive enough for
building well-known virtual machine optimization techniques, such as maps and
inline caches. We believe that they are flexible enough to express a wide
variety of language semantics efficiently.

>>>>

@@

\title{Efficiently Handling Guards in the low level design of RPython's tracing JIT}

\authorinfo{Carl Friedrich Bolz$^a$ \and David Schneider$^{a}$}
           {$^a$Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany
           }
           {XXX emails}

\conferenceinfo{VMIL'11}{}
\CopyrightYear{2012}
\crdata{}

@@

\keywords{XXX}

@@

@@

The contributions of this paper are:
\begin{itemize}
 \item
\end{itemize}

The paper is structured as follows:

@@

\subsection{RPython and the PyPy Project}
\label{sub:pypy}

The RPython language and the PyPy Project were started in 2002 with the goal of
creating a python interpreter written in a High level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative implementation of the Python language, xxx mention speed. The
Implementation takes advantage of the language features provided by RPython
such as the provided tracing just-in-time compiler described below.

RPython, the language and the toolset originally developed to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations. xxx Mention
the different language impls.

RPython is built of two components, the language and the translation toolchain
used to transform RPython programs to executable units.  The RPython language
is a statically typed object oriented high level language. The language provides
several features such as automatic memory management (aka. Garbage Collection)
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the program to a low level representations suited to be compiled
and run on one of the different supported target platforms/architectures such
as C, .NET and Java. During the transformation process
different low level aspects suited for the target environment are automatically
added to program such as (if needed) a garbage collector and with some hints
provided by the author a just-in-time compiler.

\subsection{PyPy's Meta-Tracing JIT Compilers}
\label{sub:tracing}

 * Tracing JITs
 * JIT Compiler
   * describe the tracing jit stuff in pypy
   * reference tracing the meta level paper for a high level description of what the JIT does
   * JIT Architecture
   * Explain the aspects of tracing and optimization

@@

\section{Resume Data}
\label{sec:Resume Data}

* High level handling of resumedata
   * trade-off fast tracing v/s memory usage
   * creation in the frontendÂ
   * optimization
   * compression
   * interaction with optimization
   * tracing and attaching bridges and throwing away resume data
   * compiling bridges

% section Resume Data (end)

\section{Guards in the Backend}
\label{sec:Guards in the Backend}

* Low level handling of guards
   * Fast guard checks v/s memory usage
   * memory efficient encoding of low level resume data
   * fast checks for guard conditions
   * slow bail out

% section Guards in the Backend (end)

@@

* Evaluation
   * Measure guard memory consumption and machine code size
   * Extrapolate memory consumption for guard other guard encodings
      * compare to naive variant
   * Measure how many guards survive optimization
   * Measure the of guards and how many of these ever fail

@@
